Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice

### 2020-05-31 article/main.tex
blabla section 3

\usepackage[utf8]{inputenc}
\usepackage{amssymb, amsmath, amsfonts, amscd}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{tikz}
\usepackage{minted}
\usepackage{epigraph}
\usepackage{cellspace}
\usepackage{multirow}
%\usepackage{parskip}
\usetikzlibrary{patterns}

@@

\epigraph{Any one who considers arithmetical methods of producing random digits is, of course, in a state of sin.}{John von Neumann, 1949}

%\todo{Ca, c'est du blabla...}

@@

In the realm of non-cryptographic random generators, a PRG is deemed ``good
enough'' it is passes \emph{some} efficient statistical tests --- whereas the
cryptographic notion of pseudo-randomness asks that it passes \emph{any}
efficient test. There are \textit{de facto} statistical test suites; an initial
battery of randomness tests for RNGs was suggested by Knuth in the 1969 first
edition of The Art of Computer Programming. In 1996, Knuth's tests were then
supplanted by Marsaglia's \texttt{Diehard} tests. In 2007, L'Ecuyer proposed
the \texttt{TestU01}~\cite{LEcuyerS07} library, whose ``BigCrush'' test is
considered state-of-the-art by the relevant community, to the best of our
knowledge. In 2010, the NIST proposed its own statistical test suite (Special
Publication 800-22), to which improvements were later
suggested~\cite{ZhuMLZJ16}. We understand that the \texttt{PractRand} test suite
also has a good reputation.

In any case, The goal of designers then consists in designing the simplest and
fastest possible pseudo-random generator that passes the day's favourite test
suite. A few selected ones are shown in Fig.~\ref{fig:prngs}. Deterministic
pseudo-random generators can be traced back to the work of Von Neumann

Lehmer proposed
linear congruential generators in 1949 for use on the ENIAC computer~\cite{Lehmer}

\texttt{Xorshift}~\cite{Marsaglia03},
\texttt{Xorshift128+}~\cite{Vigna17}.

\begin{figure}
  \begin{minipage}[t]{5cm}
\begin{minted}[fontsize=\small,frame=single]{C}
u128 a;
u64 m = 0xda942042e4dd58b5;

u64_t lehmer64() 
{
    a *= m;
    return a >> 64;
}
\end{minted}
  \end{minipage}
    \begin{minipage}[t]{4cm}
% 128-bit LFSR
\begin{minted}[fontsize=\small,frame=single]{C}
u32 a, b, c, d;

u32 xorshift128()
{
    u32 t = d;
    u32 s = a;
    d = c;
    c = b;
    b = s;
    t ^= t << 11;
    t ^= t >> 8;
    s ^= s >> 19;
    a = t ^ s;
    return a;
}
\end{minted}
    \end{minipage}
    \begin{minipage}[t]{4cm}
    % xorshift128+ (as in google's V8)
\begin{minted}[fontsize=\small,frame=single]{C}
u64 a, b;

u64 xorshift128plus()
{
    u64 s1 = a
    u64 s0 = b;
    a = s0;
    s1 ^= s1 << 23;        
    s1 ^= s1 >> 17;        
    s1 ^= s0;        
    s1 ^= s0 >> 26;
    b = s1;
    return a + b;
}
\end{minted}
    \end{minipage}
    \caption{Some conventional pseudo-random generators, designed for speed and simplicity.}
    \label{fig:prngs}
  \end{figure}

@@

However, problems in non-cryptographic random number generators can also have
dire consequences (barring the obvious case where they are used in lieu of their
cryptographic counterparts). When they are used in scientific computing for
Monte-Carlo simulations, their defects have the potential to alter the results
of the numerical simulation. Ferrenberg et al.~\cite{Ferrenberg92} ran a
classical Ferromagnetism Ising model Monte-Carlo simulation, in a special case
where the exact results could be computed analytically, and compared the results
of the simulation with the ``true'' answer. They used several conventional
pseudo-random generators: a 32-bit linear congruential generator, two LFSRs,
various combinations thereof, etc. They observed that changing the source of
random numbers significantly altered the outcome of the physical
simulation. Different generators produced different biases: in particular a
given LFSRs yielded energy levels that were systematically too low and critical
temperatures that were always too high, while another kind of generator yielded
the opposite (in many, repeated, trials).

@@

While the \textsf{PCG} pseudo-random generator is not meant as a cryptographic
primitive, obtaining an actual prediction algorithm requires the use of (fairly
conventional) cryptanalytic techniques. Making it practical requires in addition
a non-trivial implementation effort.

@@

When the increment is known (for instance when it is the default value), a
simplified prediction algorithm recovers the internal state from 192 bits of
pseudo-random stream. The process runs in 20 CPU minutes. It guesses 37 bits of
the internal state, then solves an instance of the \textsc{Closest Vector
  Problem} (CVP) in a 3-dimensional euclidean lattice. This requires about 50
arithmetic operations in total and reveals the entire internal state if the
guesses are correct.

@@

The LCG have been largely studied. \todo{expand this} This is why we know techniques to recover internal states knowing only upper or lower bits of the output, see \cite{Boyar1989}. A particular LCG is the Lehmer's generator \cite{Lehmer}, where the multiplier \(a\) is known and the increment is zero.

In our algorithm to predict the PCG, we need to reconstruct several truncated Lehmer's generator. For that, we will present a classical lattice-based reconstruction for a truncated Lehmer's generator, as presented in \cite{Frieze}. \todo{C'est pas exactement celle là, on dit quoi du coup ?}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amssymb, amsmath, amsfonts, amscd}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{tikz}
\usepackage{minted}
\usepackage{cellspace}
\usepackage{multirow}
%\usepackage{parskip}
\usetikzlibrary{patterns}

@@

\todo{Ca, c'est du blabla...}

@@

In the realm of non-cryptographic random generators, a PRG is deemed ``good
enough'' it is passes \emph{some} efficient statistical tests --- whereas the
cryptographic notion of pseudo-randomness asks that it passes \emph{any}
efficient test. There are \textit{de facto} statistical test suites (\todo{les
  mentionner}). The goal of designers then consists in designing the fastest
possible generator that passes the day's favourite test suite.

@@

However, problems in non-cryptographic random number generators can also have
problematic consequences (barring the obvious case where they are used in lieu
of their cryptographic counterparts). When they are used in scientific computing
for Monte-Carlo simulations, their defects have the potential to alter the
results of the simulations. Ferrenberg et al.~\cite{Ferrenberg92} ran a
classical Ferromagnetism Ising model Monte-Carlo simulation, in a special case
where the exact results could be computed analytically, and compared the results
of the simulation with the ``true'' answer. They used several conventional
pseudo-random generators: a 32-bit linear congruential generator, two LFSRs,
various combinations thereof, etc. They observed that changing the source of
random numbers significantly altered the outcome of the physical
simulation. Different generators produced different biases: in particular a
given LFSRs yielded energy levels that were systematically too low and critical
temperatures that were always too high, while another kind of generator yielded
the opposite.

@@

While the \textsf{PCG} pseudo-random generator is not meant as a cryptographic
primitive, obtaining an actual prediction algorithm requires the use of (fairly
conventional) cryptanalytic techniques.

@@

When the increment is known (for instance when it is the default value), a
simplified prediction algorithm recovers the internal state from 192 bits of
pseudo-random stream. The process runs in 20 CPU minutes (on a single core of a
server processor). It guesses 36 bits of the internal state, then solves an
instance of the \textsc{Closest Vector Problem} (CVP) in a 3-dimensional
euclidean lattice. This requires about 50 arithmetic operations in total and
reveals the entire internal state if the guesses are correct.

@@

The LCG have been largely studied, and considered as cryptographically insecure.

A reconstruction algorithm has been proposed by D. Knuth in 1985\cite{Knuth85}. The LCG he considered revealed only the upper bits of each internal state. The modulus \(2^k\) was known while the multiplier \(a\) and the increment \(c\) weren't. This algorithm could reconstruct the LCG in \(\mathcal{O}(2^l)\) where \(l\) is the number of non revealed bit of each internal state.

In 1989, J. Boyar improve the result in a first way \cite{Boyar1989}. He presented an algorithm which could predict a LCG where all the parameters (multiplier, increment and initial state) where unknown under the condition that the number of bit unrevealed was really small in comparison to the modulus.

In 1988, A. Frieze \textit{et al} \cite{Frieze} improve the result in a second way. They supposed that the multiplier \(a\) and the modulus \(2^k\) were known and used lattice-based technique to recover the LCG more efficiently.

Here we will use techniques presented in \cite{Frieze} to reconstruct some Lehmer's generator (LCG where the increment \(c\) is zero) \cite{Lehmer}.
